---
title: "R Professional Skills Course"
subtitle: "Session 3"
author: "Marc-Andrea Fiorina"
date: "2024/11/06 (updated: `r Sys.Date()`)"
output:
  xaringan::moon_reader:
    seal: false
    css: ["xaringan-themer.css", "hygge.css", "custom_css.css"]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      slideNumberFormat: "%current%"
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
library(tidyverse)
library(xaringanExtra)
```

```{r xaringan-themer, include=FALSE, warning=FALSE}
library(xaringanthemer)
style_duo_accent(
  primary_color   = "#000000",
  secondary_color = "#FFFFFF",
  link_color      = "#0645AD",
  header_font_google = google_font("Atkinson Hyperlegible"),
  text_font_google   = google_font("Atkinson Hyperlegible"),
  code_font_google   = google_font("Fira Code"),
  code_inline_background_color = "#e7e8e1",
  text_bold_color = NULL
)
```

```{r xaringan-panelset, echo=FALSE}
xaringanExtra::use_panelset()
```

class: center, middle

# Programming for Professional Research Using R

## Session 3

### April 10, 2025

---

class: middle

## Today

- Learn how to:
  - Filter, mutate, group, and summarize data using Tidyverse functions
  - Reshape data using Tidyverse functions
      
- Be introduced to:
  - The concept of a "tidy" dataset

- Practice the above!

---

class: center, middle

# Data Wrangling

---

### Tidyverse Introduction

.panelset[

.panel[.panel-name[Base R Layout]

```{r}
names(iris)
```

```{r}
str_replace(str_to_lower(names(iris)), "\\.", "_")
```

]

.panel[.panel-name[Tidyverse Layout]

Tidyverse functions introduce a 'cleaner' method to write code out, using what is called the 'pipe operator': **`%>%`**. It's almost like writing a recipe, step by step.

```{r}
names(iris)
```

```{r}
iris %>%
  names() %>%
  str_to_lower() %>%
  str_replace("\\.", "_")
```

]

]

---

### Basic Wrangling Functions

.panelset[

.panel[.panel-name[filter()]

`filter()` is used to **extract** rows (a.k.a. observations) from a dataset. It does so using a logical condition.

```{r}
filter_example <- mtcars %>%
  filter(wt > 3)

filter_example %>% head()
```

]

.panel[.panel-name[select()]

`select()` is used to **extract** columns (a.k.a variables) from a dataset. It does so using the variable(s)'s name.

```{r}
select_example <- mtcars %>%
  select(
    mpg, carb
  )
select_example %>% head()
```

]

.panel[.panel-name[mutate()]

`mutate()` can be used to either **create** a new column (a.k.a. variable) or to **modify** an existing column (a.k.a. variable).

```{r}
mutate_example <- mtcars %>%
  mutate(
    heavy = case_when(
      wt > 3 ~ "Yes",
      TRUE   ~ "No"
    )
  )
mutate_example %>% select(wt, heavy) %>% head()
```

]

]

---

### Basic Wrangling Functions

.panelset[

.panel[.panel-name[group_by() and summarize()]

`group_by()` and `summarize()` are used to **aggregate** data, i.e. to summarize information to a different level of observation.

```{r}
group_by_summarize_example <- mtcars %>%
  group_by(cyl) %>%
  summarize(
    mpg = mean(mpg, na.rm = TRUE)
  )

group_by_summarize_example
```

]

.panel[.panel-name[pivot_longer()]

```{r}
relig_income[1:6] %>% head(n = 2)
relig_income_long <- relig_income %>%
    pivot_longer(
        cols      = !religion, # Everything but religion
        names_to  = "levels",
        values_to = "num"
    )
```

]

.panel[.panel-name[pivot_longer() result]

```{r}
relig_income[1:6] %>% head(n = 2)
relig_income_long %>% head(n = 4)
```

]

.panel[.panel-name[pivot_wider()]

```{r}
fish_encounters %>% head(n = 4)
fish_encounters_wide <- fish_encounters %>%
    pivot_wider(
        names_from  = station,
        values_from = seen
    )
```

]

.panel[.panel-name[pivot_wider() result]

```{r}
fish_encounters %>% head(n = 4)
fish_encounters_wide[1:6] %>% head(n = 2)
```

]

]

---

class: center, middle

## Working with 'Tidy' Datasets

---

class: middle

```
table1
#> # A tibble: 6 x 4
#>   country      year  cases population
#>   <chr>       <int>  <int>      <int>
#> 1 Afghanistan  1999    745   19987071
#> 2 Afghanistan  2000   2666   20595360
#> 3 Brazil       1999  37737  172006362
#> 4 Brazil       2000  80488  174504898
#> 5 China        1999 212258 1272915272
#> 6 China        2000 213766 1280428583

table2
#> # A tibble: 12 x 4
#>   country      year type           count
#>   <chr>       <int> <chr>          <int>
#> 1 Afghanistan  1999 cases            745
#> 2 Afghanistan  1999 population  19987071
#> 3 Afghanistan  2000 cases           2666
#> 4 Afghanistan  2000 population  20595360
#> 5 Brazil       1999 cases          37737
#> 6 Brazil       1999 population 172006362
#> # … with 6 more rows
```

---

class: middle

```
table3
#> # A tibble: 6 x 3
#>   country      year rate             
#> * <chr>       <int> <chr>            
#> 1 Afghanistan  1999 745/19987071     
#> 2 Afghanistan  2000 2666/20595360    
#> 3 Brazil       1999 37737/172006362  
#> 4 Brazil       2000 80488/174504898  
#> 5 China        1999 212258/1272915272
#> 6 China        2000 213766/1280428583

# Spread across two tibbles
table4a  # cases
#> # A tibble: 3 x 3
#>   country     `1999` `2000`
#> * <chr>        <int>  <int>
#> 1 Afghanistan    745   2666
#> 2 Brazil       37737  80488
#> 3 China       212258 213766
table4b  # population
#> # A tibble: 3 x 3
#>   country         `1999`     `2000`
#> * <chr>            <int>      <int>
#> 1 Afghanistan   19987071   20595360
#> 2 Brazil       172006362  174504898
#> 3 China       1272915272 1280428583
```

---

class: middle

These are all useable versions of the same data. Only one of them, however, is 'tidy'.

What makes a dataset 'tidy'? From Hadley Wickham, Mine Çetinkaya-Rundel, and Garrett Grolemund, [*R for Data Science (2e)* Chapter 5 — Tidy Tidying](https://r4ds.hadley.nz/data-tidy):

1. Each **variable** must have **its own column**.
2. Each **observation** must have **its own row**.
3. Each **value** must have **its own cell**.

Easier to think about when these conditions are *not* met:

- When one variable is spread across multiple columns.

- When one observation is scattered across multiple rows.

`table1` is the tidy version of this dataset. How can we convert the other versions to be tidy?

---

.panelset[

.panel[.panel-name[`table2`]

```{r}
table2 %>% head(n = 2)

table2 %>%
  pivot_wider(
    names_from  = type,
    values_from = count
  ) %>% head(n = 4)
  
```
]

.panel[.panel-name[`table3`]

```{r}
table3 %>% head(n = 2)

table3 %>%
  mutate(
    cases      = as.numeric(str_extract(rate, ".*(?=\\/)")),
    population = as.numeric(str_extract(rate, "(?<=\\/).*"))
  ) %>% select(-rate) %>% head(n = 4)
  
```
]

.panel[.panel-name[`table4`]

```{r}
table4a %>% head(n = 2)
table4b %>% head(n = 2)
```

]

.panel[.panel-name[`table4 results`]

```{r}
table4a %>%
  pivot_longer(
    cols      = c(`1999`, `2000`),
    names_to  = "year",
    values_to = "cases"
  ) %>%
  left_join(
    table4b %>%
      pivot_longer(
        cols      = c(`1999`, `2000`),
        names_to  = "year",
        values_to = "population"
      )
  ) %>% head(n = 4)
  
```
]

]

---

class: center, middle

## Practical Exercise — Using the World Values Survey Dataset

---

<font size='+3'><b>World Values Survey</b></font>

<font size='+2'><b>Background</b></font>
<br> <br>
*"The survey, which started in 1981, seeks to use the most rigorous, high-quality research designs in each country. The WVS consists of nationally representative surveys conducted in almost 100 countries which contain almost 90 percent of the world’s population, using a common questionnaire. [...] WVS seeks to help scientists and policy makers understand changes in the beliefs, values and motivations of people throughout the world."*

<font size='+2'><b>Survey Contents</b></font>
.pull-left[
- Social values, attitudes & stereotypes
- Societal well-being
- Social capital, trust and organizational membership
- Economic values
- Corruption
- Migration
- Post-materialist index
]

.pull-right[
- Science & technology
- Religious values
- Security
- Ethical values & norms
- Political interest and political participation
- Political culture and political regimes
- Demography
]

---

class: middle

### Today's practical component

1. Successfully run the code in the `session_3.R` script

2. Create your own script and do the following:
  
    - Find mean values for 'importance in life' variables (Q1-6) for countries in another region than Europe
  
    - Calculate average 'enthusiasm' for these life subjects in countries in another region than Europe
  
    - Perform the same analysis, either on European countries or other countries, for another group of indicators in the dataset:
        - Important child qualities: Q7-18
        - Neighbors: Q19-26
        - Statements to agree with: Q27-41
        
    - Save one dataset for each of the tasks above.
    
**NOTE** — You should refer to documentation for the dataset, which can be found at [https://mfiorina.github.io/sais_r_course/](https://mfiorina.github.io/sais_r_course/)

---

class: middle

## Links

Dominic Royé, **[“A very short introduction to Tidyverse”](https://dominicroye.github.io/en/2020/a-very-short-introduction-totidyverse/)**

tidyr, **[“Pivoting”](https://tidyr.tidyverse.org/articles/pivot.html)**

Hadley Wickham, Mine Çetinkaya-Rundel & Garrett Grolemund, **[R for Data Science, 2e](https://r4ds.hadley.nz/)**

RStudio, **[RStudio Cheatsheets](https://www.rstudio.com/resources/cheatsheets/)**

